Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-03-20 paper.qmd
rewrote noise vs signal introduction

Therefore, while taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a definition and do little to help statisticians untangle and estimate the uncertainty in their projects. That being said, these taxonomies can hint towards what is important when we think about uncertainty. The location axis of the @utypo taxonomy lines up neatly with the typical data analysis pipeline and multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007] including stages as early as conceptualising the problem [@Otsuka2023] or collecting the data [@Meng2021]. The level axis from the @utypo taxonomy also hints towards an important consideration in uncertainty, because establishing how quantifiable uncertainty is informs us how it can be communicated, something that @Spiegelhalter2017 identified in the form of "precision". It is commonly noted that each source of uncertainty in an analysis must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017] (*Cite Gap 8*). It is clear that elements of these taxonomies are identified in many other comments on uncertainty and represent *real* considerations that need to be made when defining uncertainty, even if that definition is yet to be published.

@@

It is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been captured in existing uncertainty visualisation experiments. Some authors cited uncetainty was special because of its psychological qualities, but we have yet to see any evidence of that here. @Kinkeldey2014 also noticed that it is not clear whether or not uncertainty is just another variable, as many value extraction papers treats it as such in their experimental design and uncertainty may need to be in a variable class of its own. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* uncertainty information faces more difficulty. After all, confidence intervals, minimums, maxmimums, variance, and any other statistic that can be considered related to "uncertainty" is just that, a statisitc. There is nothing special about these statistics that should differentiate them from a mean or median and imply their visualisation should be "special" in a way that would warrant its own field. Especially when the core take aways from uncertainty evaluation experiments do not seem to differ at all from the established results in information visualisation. So, what is it that makes authors believe uncertainty is different?

# Are you measuing uncertainty as noise or signal?
Most evaluation experiments fail to capture uncertainty as we want to measure it, as noise. Some authors may read this and think that obviously, we dont use

Uncertainty acts as a form of statistical "hedging" for signals found in an analysis. Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020].

While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis.

This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole. It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e. when we want to communicate uncertainty as noise).

Unfortunately understanding the relevant noise seems to be its own problem, as statisticians and visualisation authors seem to be unaware or confused in how to calculate the uncertainty related to an analysis [@Hullman2020a]. Additionally, communicating uncertainty such that it conveys the appropriate singal supression involves combining uncertainty from multiple different sources, something that is considered to be difficult by even experts in uncertainty communication [@Spiegelhalter2017].

Uncertainty is noise. This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal. When we directly ask "What is the varinance of this variable" the variance *becomes* the signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it. The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

Uncertainty evaluation studies are not all value extraction. There are cases where authors *know* the uncertainty information is relevant, but are not quite sure how to capture it. These studies most commonly evaluate a visualiation on its ability to convey information for a decision or how well it induces trust. The motivation behind these experiments are obvious, if we cannot measure uncertainty *directy* maybe we can measure its secondary affects. The issue is that in order to say if some plots are *better* or *worse* than others, is to ascribe value to uncertainty information.

It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication. The conversation around uncertainty communication is often centered around the psychology is communicating risks.

In this section we will discuss how *most* attempts to measure the effects of uncertainty are either value extraction experiments with an extra step, or are measuring uncertainty using a proxy that does not actually represent uncertainty. In this section we will talk about the issues of trying to measure uncertainty, the real purpose of uncertainty visualisation, and suggestions in how we can better measure uncertainty in our visualisation experiments.

## Most experiments are value extraction experiments with extra steps
Those with a keen eye might have read the previous sections and noticed our statements were only relevant to evaluation experiments where participants must extract a particular value from a plot. Of course, these experiments are not reflective of "real world" uses of uncertainty visualisations, we do not want to visualise uncertainty so that we can answer questions such as *what is the probability of even A*. In reality uncertainty is useful for decision making and eliciting trust from an audience. This section will explain why, despite appearing to be different, these experiments are actually still value extraction experiments and thus have the same pitfalls.

## Types of Uncertainty Visualisation Papers

> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening. Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.

@@

### Decision making
Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake. The irony of an uncertainty visualisation pper expressing the importance of uncertainty visualisation, but setting up a ground truth or optimum strategy in which participants *should* ignore the uncertainty information is not lose on us. This perspective does not make sense because it implies participants should make decisions entirely based on the average. A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however it is unclear how easily participants could adopt a utility framework that is different to their own. @Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher. Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire. These papers also very rarely consider what the optimum decision would be in a given scenario, and a difficulty with anticipating the best strategy is conflated with the failings of the uncertainty visualisation. Ultimately, we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise.

@@

## Hypothesis testing versus uncertainty
It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure. Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion. All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*. It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.  There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. This increased sensisitivity could be the result of another aspect of uncertainty that is ignored. Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests. Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation? I think he did check some. I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot]. These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation. A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero). ^[Does this flip have mathematical implications in how the plots should be assessed?] This implies a visible signal in an uncertainty plot should indicate some divergence from the null.

>>>>

Therefore, while taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a definition and do little to help statisticians untangle and estimate the uncertainty in their projects. This is obvious in practice, as not understanding how to calculate uncertainty is one of the leading reasons cited by visualisation authors when explaining why they don't include it in their visualisation [@Hullman2020a]. That being said, these taxonomies can hint towards what is important when we think about uncertainty. The location axis of the @utypo taxonomy lines up neatly with the typical data analysis pipeline and multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007] including stages as early as conceptualising the problem [@Otsuka2023] or collecting the data [@Meng2021]. The level axis from the @utypo taxonomy also hints towards an important consideration in uncertainty, because establishing how quantifiable uncertainty is informs us how it can be communicated, something that @Spiegelhalter2017 identified in the form of "precision". It is commonly noted that each source of uncertainty in an analysis must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017] (*Cite Gap 8*). It is clear that elements of these taxonomies are identified in many other comments on uncertainty and represent *real* considerations that need to be made when defining uncertainty, even if that definition is yet to be published.

@@

It is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been captured in existing uncertainty visualisation experiments. Some authors cited uncetainty was special because of its psychological qualities, but we have yet to see any evidence of that here. @Kinkeldey2014 also noticed that it is not clear whether or not uncertainty is just another variable, as many value extraction papers treats it as such in their experimental design and uncertainty may need to be in a variable class of its own. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* uncertainty information faces more difficulty. After all, confidence intervals, minimums, maxmimums, variance, and any other statistic that can be considered related to "uncertainty" is just that, a statisitc. If we want to visualise the relationship between a temperature estimate and our forcasted variance on that esitmate, can we not just use a scatter plot, as we would any other pair of variables? There is nothing special about these statistics that should differentiate them from a mean or median and imply their visualisation should be "special" in a way that would warrant its own field. Especially when the core take aways from uncertainty evaluation experiments do not seem to differ at all from the established results in information visualisation. So, what is it that makes authors believe uncertainty is different?

# Uncertainty is noise, not signal
Uncertainty is noise, but most evaluation experiments measure it as signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it, if that message is invalid.  When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value. There will be noise associated with that answer and that is uncertainty. If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.

This is likely the cause of the contradiction between uncertainty visualisation papers following identical results to normal information visualisation papers, and authors believing uncertainty is inherently *different* to normal visualisation. Each of the sections in this paper likely contribute in some way to this contradiction in conceptualisation of uncertainty visualisaiton and the results of the studies in the field. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

Uncertainty, acting as it is intended, is fundamentally transparency to the viewer about a statistical analysis. There are many secondary benefits that come with this improved transparency, such as better decisions, more trust in the results and more confidence in the authors. These secondary benefits are, however, *not* uncertainty. The following sections will discuss the issues and limitations in measuring uncertainty through these secondary metrics and provide suggestions as to how future studies should consider measuring uncertainty.

## Issues with the current methods of measuring uncertainty
### Types of Uncertainty Visualisation Papers
Thankfully, we do not need to survey the literature to identify the motivations and evaluation metrics used in every uncertainty visualisation experiment, @Hullman2019 has done it for us.

L1: Behavioral Targets
- Performance (241; 64.8%): how effectively a user can extract information, make inferences, or make decisions with a visualization.
- Interpretation & Semantics (64; 17.2%): the ease with which a user associates uncertainty with an encoding.
- Quality of User Experience (67; 1.8%): the user’s valuation of the visualization, such as their preference or satisfaction.

L2: Expected Effects
Accuracy (134; 36%): Difference from a ground truth response (e.g., [37, 40]).
- Confidence/Accuracy Alignment (5; 1.3%): Correlation between confidence and accuracy (e.g., [11]).
- Impact on Decision-Making (3; 0.8%): Effects on decisions where no ground truth is implied (e.g., [15]).
- Decision Quality (14; 3.8%): Difference from a rational deci- sion standard (e.g., [23, 68]).
- Memorability (3; 0.8%): How well a fact is remembered (e.g., [39]).
- Risk Avoidance (23; 6.2%): The degree to which a user’s re- sponse attempts to avoid risk (e.g., [35]).
- Intuitiveness/Reading ease (31; 8.3%): How “naturally” a vi- sualization supports correct interpretations (e.g., [7]).
- Awareness of Judgment Process (8; 2.1%): The user’s ability to recognize aspects of their judgment (e.g., [15]).
- Learnability (0): The user’s ability to improve their perfor- mance over time
- Confidence (24; 6.5%): The degree of belief in the validity or truth of a judgment, data set, visualization, etc. (e.g., [16]).
- Satisfaction (38; 10.2%): The user’s aesthetic valuation of a vi- sualization (e.g., [22, 61]).
- Interaction (4; 1.1%): How much interaction a visualization receives in time, number of clicks, etc.
Efficiency (37; 9.9%): How well a visualization supports quick judgments (e.g., [59]).
- Understand internal model (48; 12.9%): Seeking to understand how a judgment is made (e.g. [13, 87]).

L3: Evaluation goal
- Compare impacts of multiple uncertainty visualizations (255; 68.5%): Results of one or more tasks are used to rank two or more uncertainty visualizations (e.g., [11, 40, 53]).
- Determine the impact of presenting uncertainty (98; 26.3%): At least one visualization that does not contain uncertainty infor- mation is evaluated (e.g., [68]).
- Validate effectiveness of an uncertainty vis (5; 1.3%): A user study is designed to confirm that a visualization improves some response(s). (e.g. [80]).
- Understand why/how a visualization works (54; 14.5%): Re- sponses to a visualization(s) are analyzed to identify or confirm some judgment mechanism (e.g., [87]).
- Understand interactions with user characteristics (53; 14.2%): The effect of properties of the user (expertise, location, etc.) on responses is analyzed (e.g., [1]).

L4: Measures
(A) Directly elicated measures
- Decision measures (31; 8.3%) 
  – Decision (31; 8.3%): A hypothetical choice action (e.g., saying when they would leave for the bus [23]).
  - Allocation (0%): An amount of currency wagered to en- dorse a choice (e.g., [74]).
- Attribute-based measures (141; 37.9%) 
  – Value of [nominal | ordinal | continuous] variable (67; 18%): Specifying a value along a provided scale (e.g., ask- ing the user to estimate the forecasted temperature from a display [46]).
  – Count (15; 4%): Specifying a count (e.g., of features of a certain type).
  – Probability (29; 7.8%): A probability specified as a probability or frequency (e.g., [40]) or as a Likert-style question asking for the relative probability of an event (e.g., [1]).
  – Variance (4; 1.1%): A value or set of values describing how much an outcome can vary (i.e., a range, standard de- viation, standard error, etc. as in [31]).
  – Rank by [data attribute | uncertainty attribute] (22; 5.9%): The top k features sorted by a numerical attribute.
- Self-report measures (161; 43.3%) 
  – Spontaneous Impression (16; 4.3%): An unconstrained description of one’s reaction to a visualization (e.g., [8]).
  – Subjective Confidence (28; 7.5%): A valuation of one’s confidence in information or a judgment.
  – Perceived Risk/Reliability (15; 4%): How risky, or con- versely how reliable, one believes an outcome is.
  – Perceived Ability/Effort (29; 7.8%): How hard a visual- ization is to read or use.
  – Self-reported Satisfaction (41; 10.8%): An aesthetic val- uation of a visualization.
  – Self-reported Intention (10; 2.7%): A reported intention to use or recommend information or an artifact (e.g., [17]).
  – Willingness to Pay (6; 1.6%): A reported intention in- volving allocating money to an information or artifact(s).
  – Self-reported Action (17; 4.6%): Whether one used, re- membered, etc. information or an artifact.
  - Galvanized Skin Response (GSR) (1; 0.3%): Change in the electrical resistance of the skin.
  – Interaction logs (35; 8.9%): A log of a user’s behavior as they use a visualization.
  – Eye movement (2; 0.5%): An eye tracker log of voluntary or involuntary eye movement.

(B) Derived measures

@@

### Decision making and risk aversion
Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake. The irony of an uncertainty visualisation pper expressing the importance of uncertainty visualisation, but setting up a ground truth or optimum strategy in which participants *should* ignore the uncertainty information is not lose on us. This perspective does not make sense because it implies participants should make decisions entirely based on the average. A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however it is unclear how easily participants could adopt a utility framework that is different to their own. @Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher. Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire. These papers also very rarely consider what the optimum decision would be in a given scenario, and a difficulty with anticipating the best strategy is conflated with the failings of the uncertainty visualisation. Ultimately, we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise.

@@

## Suggestions to measure uncertainty
### Heuristic checks
- Have a better hypothesis? Like its more specific. They are looking for a very specific thing *wrong* with a visualisation.
- We wary of trying to control something and observe it at the same time

### Just noticiable signal
Connecting our evaluation of uncertainty back to the reasons uncertainty is believed to be different is likely the key to improving evaluation studies. Uncertainty acts as a form of statistical "hedging" for signals found in an analysis. Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020].

### Thinking "smaller picture" with
Many visualisation experiments try to compare two plots with *several* differences, but do not seem to be interested in the *mechanisms* by which we extract information from visualisations. Let us consider a decision making experiment, there are several steps that Simply asking "Do people make better decisions with

**Experiment rant from confirmation**

### Comparisons to lineup plots and hypothesis testing
It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure. Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion. All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*. It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.  There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. This increased sensisitivity could be the result of another aspect of uncertainty that is ignored. Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests. Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation? I think he did check some. I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot]. These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation. A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero). ^[Does this flip have mathematical implications in how the plots should be assessed?] This implies a visible signal in an uncertainty plot should indicate some divergence from the null.
